Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2019-09-18 paper.tex
knitr and tex files

\documentclass{article}
\usepackage[stable]{footmisc}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[OT1]{fontenc}
\usepackage{amsthm,amsmath,amsfonts}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{xr}
\usepackage{tikz}
\usepackage{authblk}

@@

\newcommand{\yti}{Y_{Ti}}
\newcommand{\yci}{Y_{Ci}}
\newcommand{\uti}{U_{Ti}}
\newcommand{\uci}{U_{Ci}}
\newcommand{\hcont}{\bar{h}}
\newcommand{\etat}{\hcont(1)}
\newcommand{\hcontti}{\hcont(1)_{i}}

@@

<<init,include=FALSE>>=
library(knitr)
opts_chunk$set(
echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
)

@

<<runStuff,include=FALSE,eval=FALSE>>=
source('R/dataAnalysis.r')
big <- sapply(ls(),function(x) object.size(get(x))>2*1e7)
save(list=ls()[!big],file='output/analysis.RData')
@

<<loadStuff,include=FALSE>>=
load('output/analysis.RData')
@

@@

@@

Our intention is to develop, demonstrate, and contrast these three
approaches to modeling log data.
We hope that this paper can serve as a starting point for
applied educational researchers interested in mining this rich
resource, and that this paper helps spur methodological interest in
this problem.

@@

Next, as represented in Figure (b), we will conduct a mediation analysis, estimating natural
indirect effects of CTA1 treatment assignment $Z$, as mediated by
$H$. That is, we will estimate the component of $Z$'s effect on $Y$ that comes about
due to $Z$'s effect on $H$---path (1)---and $H$'s effect on $Y$, path
(2).
Notice that path (2) is identical to figure (a)---the effect of $H$ on
$Y$.
We will use this equivalence to estimate and interpret indirect
effects.
As part of the mediation analysis, we will also estimate path (3), the
natural direct effect of $Z$ on $Y$---that is, the component of $Z$'s
effect on $Y$ that is not due to hints $H$.

@@

As students work through specific problems, they receive two types
of instruction from the tutor.
When they err---that is, depart from one of the pre-specified correct
solution paths for that problem---they receive an error message
alerting them to their mistake and helping them return to the correct
solution path.
Students may also request help, in the form of hints, that are also
tailored to the specific context and solution path they are on.

@@

\begin{table}[ht]
\centering
\include{output/covariateTable}
\caption{Missingness information, control (``BaU'' or ``Business as
  Usual'') and treatment (``CTA1'') means, and balance for the
  covariates included in this study, from the high school year two
  stratum of CTA1 Effectiveness experiment. Imputation error is percent falsely classified for
  categorical variables (Race/Ethnicity, Sex, and Special Education)
  and standardized root mean squared error for Pretest, which is
  continuous. %The overall p-value accounts for clustering at the
  %school level and blocked random sampling.
  Analysis done in \texttt{R} via \texttt{RItools} \citep{ritools}.}
\label{tab:covariateBalance}
\end{table}

@@

\subsection{The Role of Hints\footnote{This subsection draws heavily
  on comments from an anonymous reviewer.}}
Students using the Cognitive Tutor Algebra software work algebra
problems on a computer, instead of on paper, as is typical.
Each problem in the software is associated with a set of skills that
the student must master, and a path (or set of paths) to solving the
problem.
As soon as a student errs in working through the problem, the
software provides feedback tailored to the specific mistake, and
designed to guide the student back to a correct solution path.
Similarly, when a student gets stuck, and is unable to determine the
next step in solving the problem, he or she can ask for a hint.
Like the error feedback, these hints are tailored to the specific
algebra skills corresponding to the next step on the solution path.
Thus, hints and error feedback are crucial aspects of the tutor's
pedagogy.

There are good reasons to believe that the availability of
hints---help on demand---may increase student learning, but there is
also reason to be skeptical \cite[See][for an overview of the theory]{aleven2016help}.
First of all, as, e.g., \citet{anderson1995cognitive} points out, ``if
students never receive help of any sort, they are in danger of becoming
permanently stuck on some problem'' (p. 190).
That is, hints give students who are stuck a way to continue working
after getting stuck.
Hints may also boost learning by ``helping students identify relevant features in
problems'' \citep[][p. 6]{aleven2016help},  that is, point students
towards aspects of the problem that are ``most important for the goals
of learning'' \citep[][p. 782]{koedinger2012knowledge}.
The final available ``bottom out'' hint typically gives the student
the answer.
Students who bottom out then see a worked example, which can be an
effective tool for learning \citep[e.g.][]{sweller1985use}.
Finally, hints can help improve students' meta-cognitive states, such
as their self-knowledge of skills.

On the other hand, all of these theoretical mechanisms rely on
students to play their part.
In order for hints to be instructive, students need to reflect on
their content, for instance with ``self-explanation''
\citep[c.f.][]{renkl1998learning}.
In the same vein, hints need to be well-designed in order to be
effective \citep{mckendree1990effective}.
This suggests that hints may be helpful for some students but not for
others.
They may even be harmful: as \citet{koedinger2007exploring} put it
(p. 241),
``many lines of research and theory suggest the importance of \dots
withholding information from students so that they can exercise, test,
or reason toward new knowledge on their own.''
\citet{paas1994variability} interpret this dilemma in terms of balancing
``cognitive load,'' that is, optimally allocating a student's
attention to the most relevant aspects of a problem.

The empirical literature on the value of hints in intelligent tutors
is mixed \citep[see][for a summary]{goldin2012learner}.
\citet{anderson1995cognitive} compared the performance of students
randomly assigned to different versions of the Cognitive Tutor with
different feedback structures, and found that students who were able
to request hints finished the lesson faster than those who were not,
but could not detect an effect on learning.
\citet{singh2011feedback} used a similar design, comparing two
versions of the ASSISTments intelligent tutor, one in which hints and
immediate error messages were available, and one which did not provide
feedback. Students in the feedback condition showed larger gains on a
posttest, though it is unclear whether these gains were due to hints
or error feedback.
On the other hand, \citet{beck2008does} and \citet{goldin2012learner}, did not randomize hint
availability, but instead used statistical models to control for
observed confounding.
Both papers used students' subsequent performance within the tutor to
estimate hint effects.
\citet{beck2008does} found conflicting results using different
analytical methods to control for confounding.
\citet{goldin2012learner} investigated two sources of heterogeneity in
the effect of a hint request: by student and by the ``level'' of the
hint (the first, second, or third ``bottom out''  hint available for a given problem).
They found that, indeed, the effect of asking for a hint varies
between students, but on average it is negative for the first two
hints requested on a problem, but positive for the bottom out hint.

\subsection{Measuring Hint Usage in the CTA1 Data}
Unlike previous studies of hint usage, the CTA1 study was not designed
to estimate the effect of hint usage; its purpose was to estimate the
effect of CTA1 as a package.
Like \citet{beck2008does} and \citet{goldin2012learner}, but unlike
\citet{anderson1995cognitive} and \citet{singh2011feedback}, hint
usage was not randomized.
On the other hand, the CTA1 study includes a posttest, like the
randomized studies but unlike the observational studies cited here.
In this regard, the CTA1 study is similar to other large-scale
evaluations of educational technology which do not randomize usage but
do include a post-test.

The CTA1 study was of a much longer duration than
any study of hint effects that we are aware of: students had access to the tutor for an entire
school year before the posttest.
This factor allows for much more data for each student than shorter
studies, as well as more variability in the types of problems and
contexts in which students may request hints.
This variability, in turn, allows researchers to fit flexible models
and ask nuanced questions about hint usage, for instance, exploring differing
patterns of usage among different students.
This sort of analysis is beyond the scope of the current study, which
focuses on comparing and contrasting causal modeling approaches for
these data.
What is relevant for our purposes is that we may narrow down the type
of hint data we model to align with our research focus and to avoid
some modeling pitfalls.

Since our posttest measured Algebra I skills, we only
considered worked problems from the Algebra I curriculum.
Next, although \citet{goldin2012learner} demonstrated the importance
of hint level, our focus is on the overall average effect of
requesting any hints; therefore, the variable measuring a student's
hint request on a problem was set one if the student requested any
hints on that problem, and zero otherwise.
(More mundanely, we do not have access to data
describing the number of hints available for every problem, which
would hinder interpretation of any other measure of hint usage.)

Lastly, any observational study of hint usage must contend with the
fact that students are unlikely to request hints on problems that they
already know how to solve (other than by accident).
Hence, any measure of hint usage is necessarily also a measure of
student ability---students who know more algebra will almost certainly
request fewer hints.
This leads to two related problems: first, student ability
confounds any statistical relationship between hint request frequency
and posttest scores.
Secondly,
a student's decision to request a hint implies both that she
found the problem at least somewhat difficult, and that, in this case,
she responded to that difficulty by requesting a hint.
Our interest is in the second implication, not the first.
Conversely, the significance of a student's decision to forgo a hint
depends on how challenging he or she finds the problem.
Hint request frequency combines students' hint requests (or non-requests) from both
problems they find challenging and trivial, so is a poor measure of
students' use of the hint feature.

Ideally, we would solve this problem by only including data on
problem-student pairs in which the problem was challenging to the
student.
Of course, no direct measure of problem challenge is available.
Instead, we considered a problem to be challenging if the student
either requested a hint or made an error on that problem.
Then, let
\begin{equation*}
\bar{h}_i=\frac{\text{\# problems on which }i\text{ requested a
    hint}}{\text{\# problems on which }i\text{ either requested a hint
    or made an error}}
\end{equation*}
measure hint frequency.

This measure is inversely related to the ratio of the percentage (of all
problems) a student makes an error without having requested a hint,
to the percentage of problems a student requests a hint
\footnote{Technically, if $h$ is the event that a student requests
a hint on a problem and $e$ is the event that the student makes
an error, then $Pr(h|h\text{ or }e)=1/\{1+Pr(h\text{ and not }e)/Pr(h)\}$}.
Making an error on a problem without having requested a hint reflects
a hesitance to request a hint when one might have been warranted.
$\bar{h}$ ignores the class of challenging problems on which students
are challenged and figure the answer out (or guess correctly) without
requesting a hint.
This class of problems is inherently interesting to our question;
however, it is unidentified.

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{h_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately \Sexpr{round(mixProp*100)} of students as
high-hint users, so we chose $c=0.6$, approximately the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$.
This definition agreed with the model's classification in
approximately \Sexpr{round(agree*100)}\% of cases.

@@

Let $Y_i$ represent subject $i$'s posttest score, and let $Z_i$ represent
$i$'s treatment status (i.e. the CTA1 or control group).
Following \citet{neyman} and \citet{rubin}, let $Y_i(Z=1)$ be
the posttest score student $i$ would exhibit were $i$ (perhaps
counterfactually) assigned to the treatment condition, and
$Y_i(Z=0)$ the score were $i$ assigned to control.
Then the CTA1 treatment effect for student $i$ is $Y_i(Z=1)-Y_i(Z=0)$.
A similar structure applies to hint request for students in the
treatment group.
For these students, let the potential outcomes be $Y_i(Z=1,H=1)$ and
$Y_i(Z=1,H=0)$ or $Y_i(H=1)$ and $Y_i(H=0)$ for short.
For a study measuring hint requests as a continuous variable, we may
define $Y_i(\hcont)$, the posttest score $i$ would exhibit were $i$ to
request hints with proclivity $\hcont$.
This structure implicitly assumes ``non-interference,'' that one
student's treatment assignment $Z$ or hint proclivity $H$ does not
affect another student's posttest scores.

Individual treatment effects are (typically) unidentified, since only
one of the relevant potential outcomes is ever observed for each
subject.
For instance, $Y(H=1)$ is unobserved for members of the
treatment group with $H=0$.
If $Pr(H_i=1)$ were known for each $i$ (as would be the case if $H$
were randomized) then the average treatment effect (ATE) of $H$ on $Y$,
$\EE[Y(H=1)]-\EE[Y(H=0)]$, could be estimated without bias or further
assumptions.
Of course, the distribution of $H_i$ is unknown, and is presumably a
function of $i$'s individual characteristics.
However, if these relevant characteristics were all measured and
available, the ATE of $H$ on $Y$ would be identified.

@@

Our solution was to dichotomize hint requests with a median split, as
described above, essentially re-defining the causal question.
Half of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the (treatment group)
median.
To make sense of the respective roles of $H$ and $\hcont$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\hcont$.
In the control group all students are assigned $H=0$ and $\hcont=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\hcont$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\hcont=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

@@

\section{Principal Stratification}\label{sec:principalStratification}
The observational study approach and mediation analysis both rely on
an ignorability assumption to estimate the effect of $H$ on $Y$, in
the latter case framing this effect as one component of the overall
treatment effect.
Principal stratification \citep[PS;][]{frangakis} takes a different approach entirely.
In the PS approach, $Y(0)=Y(Z=0)$ and $Y(1)=Y(Z=1)$ are the only potential
outcomes for $Y$.
As in mediation analysis, hint behavior has potential outcomes as
well; modeling hint requests continuously with the $\hcont$ parameter
from (\ref{eq:rasch}) gives potential values $\hcont(1)$ and $\hcont(0)$,
though only $\hcont(1)$ is relevant in the CTA1 study.

The goal of principal stratification is to estimate ``principal
effects'':
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\hcont(1)=r]
\end{equation*}
This is the treatment effect for subjects who \emph{would} request
hints at level $\hcont(1)=r$, if assigned to the treatment condition.
Whereas the focus of the observational study and mediation
analysis was the relationship between $\hcont$ and posttest scores, the
focus of PS is the relationship between $\hcont$ and treatment effects.
Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\hcont(1)=r]$,
despite the fact that $\hcont(1)$ is never observed at the same time as
$Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
strictly counterfactual quantities, as in mediation analysis.
$\hcont(1)$ is defined for subjects in both randomized groups, but only
observed in the treatment group.

In previous sections $\hcont$ was dichotomized into $H$; such
dichotomization is not necessary here.
However, $\hcont$ has some disadvantages as a continuous measure of
hint usage.
For one, since it is essentially a sample mean over a subset of each student's
worked problems, much of its variance is driven by the total number of
problems students worked---in particular, its extreme values belong to
those students who barely used the tutor at all.
Further, it does not account for varying difficulty of the tutor's
problems.
As an alternative, we fit a
modified Rasch model \citep[e.g.][]{rasch},
\begin{equation}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems in each section are equally
likely to elicit a hint.

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the larger principal stratification model, replacing $\eta$ for
$\hcont$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{output/psModel.pdf}
\caption{Results from the PS model. On the left, the estimated
  CTA1 treatment effect is plotted against hint request proclivity
  $\eta(1)$. The black line is the posterior mean, and the red lines
  are random posterior draws. The right panel plots observed posttest
  scores against one random draw of $\eta(1)$, along with regression
  lines for the treatment and control groups.}
\label{fig:psResults}
\end{figure}

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[OT1]{fontenc}
\usepackage{amsthm,amsmath,amsfonts}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{xr}
\usepackage{tikz}
\usepackage{authblk}

@@

\newcommand{\yti}{Y_{Ti}}
\newcommand{\yci}{Y_{Ci}}
\newcommand{\uti}{U_{Ti}}
\newcommand{\uci}{U_{Ci}}
\newcommand{\etat}{\eta(1)}
\newcommand{\etati}{\eta(1)_{i}}

@@

@@

The CTA1 effectiveness trial is a prominant example of a large and
growing number of randomized evaluations of educational technology
products (see \citealt{escueta2017education} for a recent systematic review).
In many, if not all, experimenters were able to
gather log data from students and/or teachers in the treatment group,
capturing the varying ways they used the technology.
These datasets could go a long way in helping researchers and
developers better understand how or for whom the software works.

@@

Our intention is to develop, demonstrate, and contrast these three
approaches to modeling log data.
We hope that this paper can serve as a starting point for
applied educational researchers interested in mining this rich
resource, and that this paper helps spur methodological interest in
this problem.
Ulitimately, we believe that education researchers should not stop at
estimating overall average treatment effects, but make use of the
wealth of log data that these RCTs produce.

@@

Next, as represented in Figure (b), we will conduct a mediation analysis, estimating natural
indirect effects of CTA1 treatment assignment $Z$, as mediated by
$H$. That is, we will estimate the component of $Z$'s effect on $Y$ that comes about
due to $Z$'s effect on $H$---path (1)---and $H$'s effect on $Y$, path
(2).
Notice that path (2) is identical to figure (a)---the effect of $H$ on
$Y$.
We will use this equivalency to estimate and interpret indirect
effects.
As part of the mediation analysis, we will also estimate path (3), the
natural direct effect of $Z$ on $Y$---that is, the component of $Z$'s
effect on $Y$ that is not due to hints $H$.

@@

As students work through specific problems, they receive two types
of instruction from the tutor.
When they err---that is, depart from one of the pre-specified correct
solution paths for that problem---they receive an error message
alerting them to their mistake and helping them return to the correct
solution path.
Students may also request help, in the form of hints, that are also
tailored to the specific context and solution path of the problem.

\subsection{The Role of Hints}\footnote{This subsection draws heavily
  on comments from an anonymous reviewer.}
Students using the Cognitive Tutor Algebra software work algebra
problems on a computer, instead of on paper, as is typical.
But the Cognitive Tutor is not simply an online repository of
problems.
Each problem in the software is associated with a set of skills that
the student must master, and a path (or set of paths) to solving the
problem.
As soon as a student errs in working through the problem, the
software provides feedback tailored to the specific mistake, and
designed to guide the student back to a correct solution path.
Similarly, when a student gets stuck, and is unable to determine the
next step in solving the problem, he or she can ask for a hint.
Like the error feedback, these hints are tailored to the specific
algebra skills corresponding to the next step on the solution path.
Thus, hints and error feedback are crucial aspects of the tutor's
pedagogy.

On the other hand, the ready availability of hints may prevent some
students from trying to figure out algebra problems themselves,
instead asking for a hint as soon as things get hard.
If hints prevent student effort, they may be detrimental---after all,
students working through traditional worksheets or textbook problems
do not have recourse to hints and may put more effort into their
schoolwork.
\citet{koedinger2007exploring} refer to this as the ``assistance
dilemma.''

The RAND CTA1 effectiveness trial provides a valuable opportunity to
study the role of hints---on average, do they tend to help students
learn or to prevent them from learning?
What role does the availability of hints play in CTA1's effectiveness
when compared to traditional instruction?
Does requesting hints make CTA1 more effective or less?

@@

\begin{table}[ht]
\centering
\include{covariateTable}
\caption{Missingness information and balance for the covariates included in this study, from
  the CTA1 Effectiveness experiment, high
  school, year two. Imputation error is percent falsely classified for
  categorical variables (Race/Ethnicity, Sex, and Special Education)
  and standardized root mean squared error for Pretest, which is
  continuous. %The overall p-value accounts for clustering at the
  %school level and blocked random sampling.
  Analysis done in \texttt{R} via \texttt{RItools} \citep{ritools}.}
\label{tab:covariateBalance}
\end{table}

@@

\section{Measuring Hint Requests}\label{sec:modelingHints}
In our CTA1 log data, 2091 students requested a total of 801,518 hints on
926,314 problems.
Some students never requested a hint, and one student requested over
three thousand hints; some students requested at least one hint on every problem
they worked.

Clearly, several factors played a role in this variation.
First, academically weaker students---those whom we would expect to
score lower on the posttest regardless of CTA1 use---probably struggle
with CTA1 problems more often, and probably request more hints.
This presents a serious confounding threat, which we will attempt to
address in a number of ways throughout the paper.
Our first line of defense is a slight restructuring of the causal
questions, reflected in the measurement stage.
By ``hint use'' we refer to a student's use of hints on problems that
he or she finds challenging.
Specifically, we consider only worked problems on which
the student either requested a hint or committed an error.
This effectively filters out the problems that each student finds
trivial; the fact that stronger students struggle less often is no
longer a driving factor.
(In an equivalent formulation, the question addresses each student's
likelihood of committing an error without having asked for a
hint---i.e. deciding to attempt a difficult problem without help.)
This strategy removes some of the confounding between
student ability and hint requests, though (as we shall see) a
substantial amount remains.

Factors beyond student ability also play a role in the variation
in the numbers of hints students requested.
For one, students who worked a larger number of CTA1 problems also
requested more hints.
Secondly, not all problems are equivalent.
Problems vary in their structure, and hence in the number of hints
available---some problems allow an indefinite number of hint requests.
Problems from some sections never elicited hints, whereas problems
from other sections nearly always did.
A students' hint requests are a function of their learning
style within the tutoring software, which parts of the software they
worked, and how much they worked.

A focus on hint requests, as opposed to these other
factors, requires some care.
Modeling hint requests per problem as binary---did a student request
at least one hint?---puts problems with many or few
available hints on equal footing.
Problem difficulty is another matter.
What is difficult for one student may not be difficult for others.
That students do not request hints on problems they find trivial is
uninteresting; what is more interesting is how they react when a
problem presents a challenge.

\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{rasch.jpg}
\caption{The proportion of problems in which each student requested a
  hint, or the IRT student parameter $\eta$ from model
  (\ref{eq:rasch}), as a function of each student's average section difficulty. For each
  section, ``section difficulty'' is defined as the proportion of
  \emph{that section's} worked problems that elicited a hint request,
  and a student's average section difficulty is the average section
  difficulty over that student's worked problems.}
\label{fig:mbar}
\end{figure}

To capture the effect of hint requesting behavior, as distinct from
total amount of usage, a first step may be to calculate the proportion
of each student's challenging problems that resulted in a hint
request.
However, this presents additional issues, one of which can be seen in Figure
\ref{fig:mbar}.
First, even after dichotomizing hints and restricting our focus to
instances in which students appear to be challenged, worked problems
may differ from each other in relevant ways.
For instance, multi-step problems which provide more occasions for hint
requests may elicit at least one hint more frequently.
For each section of the tutor, let ``section difficulty'' be the proportion of that
section's challenging problems that elicited a hint request; let a
student's ``average section difficulty'' be the average section
difficulty over all challenging problems that student worked.
On the left of Figure \ref{fig:mbar}, the proportion of problems on
which each student requested at least one hint is plotted against that
student's average section difficulty.
It is apparent, if unsurprising, that students who worked on sections of the
curriculum that typically elicited more hints tended to request hints
more often.
The Spearman correlation between average section difficulty and hint
request proportion is 0.35.
How do we equate the hint records of students who worked problems in different
sections?

% Further, hint request proportions are related to the number of
% challenging problems each student worked, as can be seen on the right
% panel of Figure \ref{fig:mbar}.
% This is due to measurement error.
% The variance of a sample proportion is inversely proportional to the
% size of the sample; therefore, it is unsurprising that the extremes of
% hint request proportions are composed primarily of students who worked
% very few problems.
% For instance, nearly all students who requested a hint on more than
% 75\% of the problems that challenged them worked fewer than 250
% challenging problems.

Item response theory \citep[e.g.][]{irtHandbook} provides an attractive alternative.
Under a modified Rasch model \citep[e.g.][]{rasch}, for instance,
\begin{equation}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems in each section are equally
likely to elicit a hint.
Replacing student hint request proportions with the posterior mode for
$\eta$ from model (\ref{eq:rasch}), fit using the \texttt{lme4}
package in \texttt{R} \citep{lme4}, resolves the section-difficulty conundrum: the right
side of Figure \ref{fig:mbar}, in which $\eta$ is plotted instead of
hint request proportion, shows no substantial relationship between
section difficulty and hint requests.

In some analyses, we dichotomize $\eta$ with a median split.
Let $H_i=1$ if $\eta_i>median(\bm{\eta})$ and 0 otherwise.
While in dichotomizing $\eta$ we lose valuable information, doing so
simplifies the observational matching design of the following section,
and makes the mediation analysis possible.

@@

Let $Y_i$ represent subject $i$'s posttest score, and let $Z_i$ represent
$i$'s treatment status (i.e. the CTA1 or control group).
Following \citet{neyman} and \citet{rubin}, let $Y_i(Z=1)$ be
the posttest score student $i$ would exhibit were $i$ (perhaps
counterfactually) assigned to the treatment condition, and
$Y_i(Z=0)$ the score were $i$ assigned to control.
Then the CTA1 treatment effect for student $i$ is $Y_i(Z=1)-Y_i(Z=0)$.
A similar structure applies to hint request for students in the
treatment group.
For these students, let the potential outcomes be $Y_i(Z=1,H=1)$ and
$Y_i(Z=1,H=0)$ or $Y_i(H=1)$ and $Y_i(H=0)$ for short.
For a study measuring hint requests as a continuous variable, we may
define $Y_i(\eta)$, the posttest score $i$ would exhibit were $i$ to
request hints with proclivity $\eta$.
This structure implicitly assumes ``non-interference,'' that one
student's treatment assignment $Z$ or hint proclivity $H$ does not
affect another student's posttest scores.

Individual treatment effects are (typically) unidentified, since only
one of the relevant potential outcomes is ever observed for each
subject.
For instance, $Y(H=1)$ is unobserved for members of the
treatment group with $H=0$.
If $Pr(H_i=1)$ were known for each $i$ (as would be the case if $H$
were randomized) then the average treatment effect (ATE) of $H$ on $Y$,
$\EE[Y(H=1)]-\EE[Y(H=0)]$, could be estimated without bias or further
assumptions.
Of course, the distribution of $H_i$ is unknown, and is presumeably a
function of $i$'s individual characteristics.
However, if these relevant characteristics were all measured and
available, the ATE of $H$ on $Y$ would be identified.

@@

Our solution was to dichotomize hint requests with a median split, as
described above, essentially re-defining the causal question.
Half of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the (treatment group)
median.
To make sense of the respective roles of $H$ and $\eta$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\eta$.
In the control group all students are assigned $H=0$ and $\eta=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\eta$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\eta=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

@@

\section{Principal Stratification}\label{sec:principalStratification}
The observational study approach and mediation analysis both rely on
an ignorability assumption to estimate the effect of $H$ on $Y$, in
the latter case framing this effect as one component of the overall
treatment effect.
Principal stratification \citep[PS;][]{frangakis} takes a different approach entirely.
In the PS approach, $Y(0)=Y(Z=0)$ and $Y(1)=Y(Z=1)$ are the only potential
outcomes for $Y$.
As in mediation analysis, hint behavior has potential outcomes as
well; modeling hint requests continuously with the $\eta$ parameter
from (\ref{eq:rasch}) gives potential values $\eta(1)$ and $\eta(0)$,
though only $\eta(1)$ is relevant in the CTA1 study.

The goal of principal stratification is to estimate ``principal
effects'':
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
This is the treatment effect for subjects who \emph{would} request
hints at level $\eta(1)=r$, if assigned to the treatment condition.
Whereas the focus of the observational study and mediation
analysis was the relationship between $\eta$ and posttest scores, the
focus of PS is the relationship between $\eta$ and treatment effects.
Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\eta(1)=r]$,
despite the fact that $\eta(1)$ is never observed at the same time as
$Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
strictly counterfactual quantities, as in mediation analysis.
$\eta(1)$ is defined for subjects in both randomized groups, but only
observed in the treatment group.

In typical PS studies, the intermediate variable is modeled as fixed
and known.
We chose, instead, to measure hint behavior as a latent variable from
Rasch model (\ref{eq:rasch}), due to the reasons described in Section
\ref{sec:modelingHints} and because of the measurement error
due to the varying number of problems each subject worked.
A full treatment of this distinction and its implications, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{psModel.pdf}
\caption{Results from the PS model. On the left, the estimated
  CTA1 treatment effect is plotted against hint request proclivity
  $\eta(1)$. The black line is the posterior mean, and the red lines
  are random posterior draws. The right panel plots observed posttest
  scores against one random draw of $\eta(1)$, along with regression
  lines for the treatment and control groups.}
\label{fig:psResults}
\end{figure}
